---
title: "Trends in Toronto Public Libraries: A Study of 2012-2022 Toronto Public Library Data"
subtitle: "My subtitle if needed"
author: Liam Wall
thanks: "Code and data are available at: https://github.com/Lwall02/What-Are-We-Using-Libraries-For."
date: today
date-format: long
abstract: "This paper looks into four annual statistics released by the Toronto Public Library (TPL) system from 2012-2022 - visits, new registrations, circulation of materials, and computer sessions usage. It shows an almost predictable down-trend in library from 2012 to 2019 and a major hit to the TPL system during the COVID-19 pandemic. Follwowing the 2020 hit, the TPL began to bounceback but all of the statistics (bar perhaps one) do not recover to pre-pandemic levels. The only positive correlation in the entire paper is between new registrations and computer usage from 2012-2015."
format: pdf
toc: true
number-sections: true
bibliography: references.bib
---

```{r}
#| include: false
#| warning: false
#| message: false

library(tidyverse)
library(knitr)
library(patchwork)

clean_data <- read_csv("/cloud/project/outputs/data/cleaned_combined_data.csv")


```


# Introduction

With the ever-increasing accessibility of publications, like books, magazines, and newspapers, to name a few, one can wonder what libraries are really used for nowadays. It is well-known that a library is a place to come to read and learn. There is an immensity of resources available at most public libraries across all of the world, including public records, films, local newspapers, computers, socializing areas, and of course publications. Despite the common thought that not many people go to libraries, it is still a strong belief that the absence of libraries would be a detriment to the surrounding community [@pewresearch]. 

To a lot of people, the ability to access the internet is nothing more than sitting at their personal home computer or using their phone connected to their wifi or cellular service. However, one does not always remember there is a significant portion of the community that is not always fortunate enough to have this level of access. In today's world, this can be one the libraries single most important amenities; access to the internet. The Toronto Public Library (TPL) system offers free wifi at every branch and almost all of their 103 branches offer free computer stations [@torontopubliclibrary]. In this paper, I look at TPL information from 2012 to 2022 covering the number of annual visits, new registrations, circulation of materials (called circulations), and computer station usage (called a computer sessions). I wanted to find out why people are using the TPL, when it was most popular, what is increasing or decreasing, and how popular are libraries today compared to a decade ago.

My findings show that there is a definitive decrease in every statistic from 2012 to 2022 except for new registrations. They show a decrease in yearly visitors, circulations, and computer sessions, however the number of new registrations per year has countered these trends, in fact the number of new registrations in 2022 is more than that of 2012. We can also see the large impact the COVID-19 pandemic had on the TPL with decreases in every statistic during 2020. Post-pandemic, the numbers in each of the four categories began to increase however, only new registrations were able to reach pre-pandemic levels by 2022.

The Data section of this paper provides an explanation on the TPL data used in this paper and the steps that went into making this data usable and readable for a report. The Results sections discusses in greater detail the findings from the data by showing the trends I explained. The Discussion sections provides some information on the possible shortcomings of the analysis and other events that may have an effect on the findings. The Conclusion sections summarizes the findings from this paper. 

\newpage

# Data

Data used in this paper was downloaded from the Open Data Toronto website [@opendatatoronto][@librarydatasets]. I downloaded a total of five data sets published by the Toronto Public Library: Library Branch General Information [@geninfo], Library Visitors [@visits], Library Workstation Usage [@sessions], Library Card Registrations [@regis], and Library Circulation [@circ]. The data was worked with entirely on the open source programming language R [@citeR]. I used the following packages to clean and analyze the data: tidyverse [@tidyverse], ggplot2 [@ggplot], patchwork [@patchwork], knitr [@knitr], and dplyr [@dplyr].

## The Data Sets 

I downloaded the five above mentioned data sets and in order to conduct some analysis I had to merge four of them into one tibble. The cleaning and merging of the data sets is explained below.

It is important to note that there were 100 to 104 different TPL branches across the Greater Toronto Area each year from 2012 to 2022. Each branch is attached with a branch code that is two capital letters. For example, the Lillian H. Smith Library located just below the University of Toronto is labeled LS. You will see these branch codes in tables and graphs later in the paper.

### The Library Branch General Information

The Library Branch General Information data set was only used in two ways for this paper. It provided a key to attach a libraries branch code with its full name and address, and it separated the physical branches from the 'non-physical' branches. The physical branches are the libraries you can visit, the 'non-physical' libraries are services like a bookmobile or a virtual library. The 'non-physical' libraries are shown in @tbl-nonphysicalbranches. (SHOW AND REFERNCE TABLE OF THE NONPHYSICAL LIBRARIES). 

```{r}
#| label: tbl-nonphysicalbranches
#| tbl-cap: "This shows the 12 non-physical branches as of 2023 along with their TPL given name"
#| tbl-cap-location: bottom
#| echo: false
#| message: false
#| warning: false

raw_gen_info <- read_csv("/cloud/project/inputs/data/tpl-branch-general-information-2023.csv")

raw_gen_info |>
  filter(PhysicalBranch == 0) |>
  select(BranchCode, BranchName, PhysicalBranch) |>
  kable(col.names = c("Branch Code", "Branch Name", "Physical Branch (1 - Yes/ 0 - No)"),
        booktabs = TRUE,
        linesep = "")
```


It is important to separate and remove the 'non-physical' libraries from all the data sets because they introduce very different values in each of the categories I am looking at. The 'non-physical' libraries recorded their visits, registrations, circulations, and computer sessions in different ways. For example, the visits would have been clicks to a website or number of visitors into a mobile library. In this paper, I am focused on the popularity of the physical library and so this data does not serve to tell me how a physical library is performing.  

### The Library Card Registration, Circulation, Visits, and Workstation Usage

These four data sets, The Library Card Registration, Circulation, Visits, and Workstation Usage, serve as the essential data for this paper. As they are also released by the TPL system, they follow the exact same branch code organization. Each of the four data sets gives data for each branch code for every year from 2012 to 2022. They are very similar data sets to each other in that there is the year, branch code, and number of the value that is to be shown (visits, registrations, circulations, or sessions).

## Cleaning the Data

The first step of cleaning the data so that I could analyse it in an efficient way was to combine the visits, registrations, circulations, and sessions, into one table according the year and branch code. In that way all the information stays according to the correct library it was taken from and the correct year. 
The second step was to use the clean_name() function. The third step was to remove the 'non-physical' branches so that the data I analyse is from real, physical libraries only. A sample of final table with each of the four statistics aligned with branch code and year is found in @tbl-cleandatatable. (SHOW A TABLE OF THE COMPLETE INFO I USED). An explanation of the four variables is found in Results.

```{r}
#| label: tbl-cleandatatable
#| tbl-cap: "This shows 10 rows of the final table used for analysis. It contains every branches information for each year from 2012-2022 organized such that the most visited library per year is at the top"
#| tbl-cap-location: bottom
#| echo: false
#| message: false
#| warning: false
clean_data |>
  arrange(year, -visits) |>
  slice(1:10) |>
  kable(col.names = c("Year", "Branch Code", "Visits", "Registrations", "Circulations", "Sessions"),
        booktabs = TRUE,
        linesep = "")
```


# Results

First, before I get into the results, I will explain what each variable is that you will see in this section. The 'Visitors' describes the number of visitors per year at a certain library branch. The 'Registrations' describes the number of library cards registered at a certain library branch per year. The 'Circulation' describes the number of materials that were checked out and returned during the year at a certain library branch. The 'Sessions' describes the number of computer sessions at a library's public workstation per year. A computer session lasts a duration of at least 15 seconds.

Second, the number of library branches did not stay the same every year. There were anywhere from 100 to 104 library branches, as mentioned in the introduction, however there was only ever 96 to 100 active physical branches. This can be seen in @fig-numbranches.

```{r}
#| label: fig-numbranches
#| fig-cap: "This shows the variation in the number of library branches each year"
#| fig-cap-location: bottom
#| echo: false
#| warning: false


clean_data |>
  count(year) |>
  ggplot(aes(x = year, y = n)) + 
  geom_line(linetype = "dashed") +
  geom_point() +
  theme_classic() +
  labs(title = "Number of Active Toronto Public Libraries") +
  xlab("Year") +
  ylab("Number Open") + 
  xlim(2012, 2023)
```

\newpage

The results of this analysis followed from looking at the trends of the data over the years. The first step was to try and see how the libraries performed each year based on the four variables listed above. The four scatter plots in @fig-scattertotaldata show the count of each variable from 2012 - 2022. In this format it is hard to see whether visits are increasing or decreasing, especially from 2012 - 2019, and the same goes for the other three variables. The biggest takeaway from this view is that the COVID-19 pandemic did effect the libraries in 2020. It is hard to say exactly how much the libraries were effected but we can see that the visits, registrations, circulations, and sessions, have no very large values in 2020. 

```{r}
#| label: fig-scattertotaldata
#| fig-cap: "A visual of all the data used for this analysis, organized by year. Each color represents a value for a library branch"
#| fig-align: 'center'
#| echo: false
#| warning: false

sp1 <- clean_data |> 
  ggplot(aes(x = year, y = visits, color = branch_code )) +
  geom_jitter(alpha = 0.3, width = 0.5, height = 0.5) +
  theme_minimal() +
  labs(x = "Year", y = "Annual Visits", color = "Branchcode") +
  xlim(2011, 2023) +
  theme(legend.position = "none") 

sp2 <- clean_data |>
  ggplot(aes(x = year, y = registrations, color = branch_code )) +
  geom_jitter(alpha = 0.3, width = 0.5, height = 0.5) +
  theme_minimal() +
  labs(x = "Year", y = "New Registrations", color = "Branchcode") +
  xlim(2011, 2023) +
  theme(legend.position = "none") 

sp3 <- clean_data |>
  ggplot(aes(x = year, y = circulation, color = branch_code )) +
  geom_jitter(alpha = 0.3, width = 0.5, height = 0.5) +
  theme_minimal() +
  labs(x = "Year", y = "Circulations", color = "Branchcode") +
  xlim(2011, 2023) +
  theme(legend.position = "none") 

sp4 <- clean_data |>
  ggplot(aes(x = year, y = sessions, color = branch_code )) +
  geom_jitter(alpha = 0.3, width = 0.5, height = 0.5) +
  theme_minimal() +
  labs(x = "Year", y = "Computer Sessions", color = "Branchcode") +
  xlim(2011, 2023) +
  theme(legend.position = "none") 

sp1 + sp2 + sp3 + sp4
```

\newpage

In @fig-scattertotaldata, it is hard to determine any kind of trend among the visits, registrations, circulations, and computer sessions. As stated above, it especially hard to do so from 2012 - 2019 because there is so many points crowding the same space. In order to look at the trends of this data, we look at the averages of each variable. This is visible in @fig-averagedata.


```{r}
#| label: fig-averagedata
#| fig-cap: "A visual of the average of each variable from 2012 - 2022. We can more clearly see the negative trends of each variable."
#| echo: false
#| message: false
#| warning: false

summarised_data <- clean_data |>
  drop_na() |>
  summarise(
    mean_visits = mean(visits),
    mean_registrations = mean(registrations),
    mean_circulation = mean(circulation),
    mean_sessions = mean(sessions), 
    .by = year
  )

plot_mean_visits <- summarised_data |>
  ggplot(aes(x = year, y = mean_visits)) +
  geom_point(alpha = 0.5, color = "red") + # geom_jitter(alpha = 0.5, width = 1, height = 1) +
  theme_bw() +
  labs(x = "Year", y = "Annual Visits") +
  xlim(2011, 2023) +
  theme(legend.position = "none") 

plot_mean_regis <- summarised_data |>
  ggplot(aes(x = year, y = mean_registrations)) +
  geom_point(alpha = 0.5, color = "blue") + # geom_jitter(alpha = 0.5, width = 1, height = 1) +
  theme_bw() +
  labs(x = "Year", y = "New Registrations") +
  xlim(2011, 2023) +
  theme(legend.position = "none")

plot_mean_circ <- summarised_data |>
  ggplot(aes(x = year, y = mean_circulation)) +
  geom_point(alpha = 0.5, color = "purple") + # geom_jitter(alpha = 0.5, width = 1, height = 1) +
  theme_bw() +
  labs(x = "Year", y = "Circulations") +
  xlim(2011, 2023) +
  theme(legend.position = "none")

plot_mean_sessions <- summarised_data |>
  ggplot(aes(x = year, y = mean_sessions)) +
  geom_point(alpha = 0.5, color = "orange") + # geom_jitter(alpha = 0.5, width = 1, height = 1) +
  theme_bw() +
  labs(x = "Year", y = "Computer Sessions") +
  xlim(2011, 2023) +
  theme(legend.position = "none")
  
plot_mean_visits + plot_mean_regis + plot_mean_circ + plot_mean_sessions
```
As you can more clearly see in @fig-averagedata, yearly visits has been steadily decreasing across all TPL. Further, yearly circulations has really been declining since 2012. When looking at the new registrations and computer usage there is more storied trend. Surprisingly, despite less visitors and less materials being checked out, computer sessions and new registrations were increasing from 2012 - 2015. From 2016 onwards, computer sessions have trended down quite significantly while new registrations has remained trending upwards. 

Once COVID-19 struck in 2020, every variable took quite a hit. In @fig-averagedata, we can see every variable has gone near the x-axis at 2020. We can also see that from 2020 - 2022, visitors, circulations, and especially new registrations have trended upwards. In fact, new registrations is averaging the same amount in 2022 as it was for nearly all of 2012 - 2019. Average computer sessions, however, has remained very low compared to its previous averages. 

\newpage

# Discussion

After looking at this data, we can say that the libraries are not being used as much as they were in 2012. The average number visitors per year was slowly decreasing from 2012 - 2019, and post-pandemic these numbers have not recovered to where they should have been if it followed this trend in 2020. This can be due to many things, but the biggest and most obvious of those being people are reading less. We can see this clearly in the @fig-averagedata where the trend of circulations has been steeply decreasing across 2012 - 2019. In 2012 there was an average of around 310,000 circulations per year where in 2019 there was an average of around 220,000 circulations. That is to say circulations have decreased almost 30% in 2019 compared to 2012. 

As for the new registrations and computer sessions we see a somewhat different looking trend. First of all, computer sessions was trending upwards until 2015, however, after 2015 the average number of computer sessions began quickly declining. In 2018 and 2019, computer sessions took an even larger decrease in average yearly use. On the other hand, new registrations has been slowly, but steadily, increasing from 2012 - 2019. In contrast to the computer sessions, new registrations were the highest they have ever been in 2018 and 2019. And post-pandemic we can see that new registrations have risen to meet their pre-pandemic average where average yearly computer sessions have remained quite low in 2020 - 2022.

Of the four statistics, new registrations is the only one with a steady increase, and the only one especially to meet its pre-pandemic averages. Visitors, circulations, and computer sessions have all remained relatively down. Among these, circulations is the only one with a steady increase since 2020. Visitors have increased greatly since 2022, surpassing the lowest yearly averages in 2020 and 2021 but they still remain about 50% of what they used to be pre-pandemic. 

It is important to acknowledge that this report is not without flaws and limitations. In particular, I can point to a limit being that there was some data lost in while merging the library data sets. There is another limitation in that the averages could only be taken over the libraries that had data in the corredt column. That is some library branches had no data in certain years for circulation, registration, and computer sessions. Also, working with strictly annual data may reduced the accurateness of some of the statistics, particularly during 2020-2021 where monthly data may have provided a better estimate on the performance of the libraries in these years. Perhaps some of the libraries were closed by the government during certain months and annual data does not reflect this.


# Conclusion

Overall, the Toronto Public Library system is in worse shape than in 2012. This partly due to the fact many people have access to the internet where they no longer require getting physical copies of books or even going to the library to use the internet. And another part in that the TPL didn't act quick enough to counter the decreasing visitors and decreasing circulation. There is something to be said for the large increase in new registrations during 2018 and 2019, but because of the pandemic we won't be able to know if this would've increased the other variables averages. 

By 2022, every variable had experienced increases since 2020, so in a couple years time there could be major increases that would counter the argument of this paper. And to sum up the argument of this paper, libraries need to adapt to today's trends because right now they are still losing popularity. Nowadays people are reading less and have much easier access to the internet than even in 2015. The greatest successes of the TPL was in 2012 - 2015 where despite major circulation decreases, computer usage and registration were, in fact, increasing. This is almost ten years ago so unfortunately, the pandemic can not explain all of the TPL's loss in popularity.   

\newpage
# References